---
title: "Covid91 vaccine study Final2022"
author: "Benjamin Hardani"
output: 
  html_document:
    number_sections: true
    toc: true
    toc_depth: 5
date: "Summer 2022" 
---
 
```{r include=FALSE,echo=FALSE}
require(tidyverse)
require(tigerstats)
require(forcats)
FinalDat <-read.csv(file="final_Data.csv",header=TRUE)
#recode to remove the annoying "X" from the dataframe
FinalData <- FinalDat  %>% select(-one_of("X"))
FinalData <- FinalData %>%
mutate(across(where(is.character), as.factor))
Male = filter(FinalData, sex =='M')
Female = filter(FinalData, sex =='F')
LGBTQ = filter(FinalData, LGBTQ =='gay')
Druggie = filter(FinalData, DrugUser =='yes')
```

# Introduction

A new  preventative vaccine (covid91) is under test in clinical trials. Given the nature of this vaccine it is suspected that the drug may exhibit different effectiveness across the variety of subpopulations tracked. Consider population subgroups by gender( guys and gals), drug use, and sexual orientation. Write a complete report that separately considers this drug for each subgroup and your recommendations as to whether people in the subgroup should take the vaccine. Structure your report in four sections each devoted to one of the four subgroups, use null and alternative hypothesis statements and correct statistical verbage when referencing p-values in conclusions.

### Hypotheses

$H_0$ = The effectiveness of the vaccine does not change between the 4 groups of people.

$H_A$ = The effectiveness of the vaccine does change between the 4 groups of people.

### Methods

In order to determine the relation between the different subgroups and the Covid-19 vaccine, I will assess each subgroup individually. This is a Cat~CAT situation since both variables (Subgroup and the Effectiveness of the vaccine.) For each individual subgroup, I will analyze the Graphical descriptive results via a bar chart, the numerical descriptive results with a table, the inferential results with a chi-squared test, and the inferential descriptive results through a Fischer exact test. With this data, I will determine the effectiveness of the vaccine and state my recommendation for each subgroup's use of the vaccine.


## Males

### Graphical Descriptive Results

```{r}
barchartGC(~infected + treatment,data=Male)
barchartGC(~infected + treatment,data=Male, type="percent")
```

Chart No. 1 compares the effectiveness of the placebo and the vaccination in preventing covid in the men subpopulation. The percentages from the first graph are depicted in the second graph in respect to men. The graph demonstrates that the men's vaccination rates only slightly improved. Compared to the placebo group, the males who received the vaccine had an immunity to the virus of about 35%. In the "OK" area, men and women are equally divided in both graphs. The number of instances of COVID in men appears to be declining thanks to the vaccine.

### Numerical Descriptive Results

```{r}
table1 <- xtabs(~infected + treatment, data=Male)
rowPerc(table1)
colPerc(table1)
```

The percentages of people who had COVID 19 are shown in this table. When given a placebo, almost 68 percent of males actually developed COVID, compared to about 32 percent of those who received the vaccine. Anyone who received the vaccine had a 36% lower risk of contracting the virus.

### Inferrential Results

```{r}
chisq.test(table1)
chisqtestGC(table1)
fisher.test(table1)
```

 The Fisher exact test gave an odds Ratio of 2.092 and has a 95% Confidence interval from 1.667079 to 2.638055.Since the odds ratio is above 1, that means the vaccine is effective in preventing cases of covid. This number means that the vaccine is approximately 2.1 times as effective as the placebo for the male subjects for preventing covid. The chi-squared test has a p-value is 2.645e-11, which is less than .05.We can reject the null hypothesis of "The effectiveness of the vaccine does not change between the 4 groups of people.".

### Efficacy

```{r}
efficMale = 2.092161/(2.092161 + 1) * 100
efficMale = round(efficMale)
confIntLower = 1.667079/(1.667079+ 1) * 100
confIntLower = round(confIntLower)
confIntUpper = 2.638055/(2.638055 + 1) * 100
confIntUpper = round(confIntUpper)
```

The efficacy for males was `r efficMale`%.  
The 95% Confidence Interval for efficacy was from (`r confIntLower`, `r confIntUpper`)%.

### Male Conclusion

 As seen in the Fisher exact test, with a p-value of less than .05, we rejected the null hypothesis of"The effectiveness of the vaccine does not change between the 4 groups of people." The bar chart and the data both show that Men who received the  vaccine were around 35% less likely to catch covid. The conclusion can be drawn that it is safe and efective for men to get the covid vaccine.



## Females

### Graphical Descriptive Results

```{r}
barchartGC(~infected + treatment,data=Female)
barchartGC(~infected + treatment,data=Female, type="percent")
```

Because females who receive the placebo are roughly 18% more likely to contract COVID than those who receive the vaccination, the Covid shot appears to be effective for protecting against COVID in females. The graph on the right displays the percentages of the left graph while the graph on the left displays the total number of people who were tested. Under the section "OK," both graphs exhibit an equal distribution of male and female respondents. For women, the Covid shot appears to be effective.

### Numerical Descriptive Results

```{r}
table2 <- xtabs(~infected + treatment, data=Female)
rowPerc(table2)
colPerc(table2)
```

The percentages for those who received COVID 19 are shown in this table. Only 41% of women who received the vaccine contracted covid, compared to about 59 percent of women who received the placebo treatment. Those who received the vaccination had an 18% lower risk of getting the disease. Although the vaccine worked, ladies had a 9 percent lower success rate than their male counterparts.

### Inferrential Results

```{r}
chisq.test(table2)
chisqtestGC(table2)
ftestF = fisher.test(table2)
ftestF
effecF = ftestF$estimate/(ftestF$estimate + 1) * 100
```

The Fisher exact test gave an odds Ratio of 1.428903 and a 95% Confidence interval from 1.150398 to 1.778592. The odds ration is above 1, so the vaccine is effective in preventing covid in females. This means that the Covid vaccine is about 1.4 times more effective for females then the placebo group, which is still effective, but less than the Males. The p value is 0.0009484 < .05; therefor, we can reject the null hypothesis. Therefore we can conclude it is  effective.

### Efficacy

```{r}
efficFemale = 1.428903/(1.428903 + 1) * 100
efficFemale = round(efficFemale)
confIntLower = 1.150398/(1.150398+ 1) * 100
confIntLower = round(confIntLower)
confIntUpper = 1.778592/(1.778592 + 1) * 100
confIntUpper = round(confIntUpper)
```

The 95% Confidence Interval for efficacy is (`r confIntLower`, `r confIntUpper`)%.
The efficacy for Females is `r effecF`%. 


### Female Conclusion

The difference in covid instances between the placebo group and the vaccine group in the male research was roughly 36%, whereas it was exactly 50% in the female study. Women who receive the vaccine are more protected from Covid than women who receive a placebo. Despite the vaccine's lower effectiveness in females, getting it is still a recommended idea. Long-term safety and efficacy are intact, thus it should be used.

## LGBTQ

### Graphical Descriptive Results

```{r}
barchartGC(~infected + treatment,data=LGBTQ)
barchartGC(~infected + treatment,data=LGBTQ, type="percent")
```

The graphs demonstrate that vaccine recipients experienced considerably more incidences of COVID than placebo recipients. Additionally, the section "OK" is now 56/44 rather than 50/50 as it formerly was. LGBTQ individuals had a roughly 70% increased risk of developing covid compared to those receiving a placebo. 

### Numerical Descriptive Results

```{r}
table3 <- xtabs(~infected + treatment, data=LGBTQ)
rowPerc(table3)
colPerc(table3)
```

With the number of covid instances for the placebo group being merely 13.33 percent, and the vaccine group at 86.67, these results are significantly different from those of the preceding two groups, Men and Women. This vaccine is less recommended than the first two groups since those who receive it are 73.34 percent more likely to develop Covid than those who receive a placebo. 

### Inferrential Results

```{r}
chisq.test(table3)
chisqtestGC(table3)
fisher.test(table3)
```

The Fisher exact test gave an odds Ratio of 0.1465164 and a 95% Confidence interval from  0.05991122 to 0.31196422. Since the odds ration is below 1, meaning the vaccine is not at all effective in preventing the covid virus. This means that the Covid vaccine is on average .15 times as effective as the placebo for members in the LGBTQ comunity, which is the lowest value thus far. The p-value is 1.642e-09, which is less than .05, meaning we must, like the last 2 times, reject the null hypothesis.

### Efficacy

```{r}
efficLGBTQ = 0.1465164/(0.1465164 + 1) * 100
efficLGBTQ = round(efficLGBTQ)
confIntLower = 0.05991122/(0.05991122+ 1) * 100
confIntLower = round(confIntLower)
confIntUpper = 0.31196422/(0.31196422 + 1) * 100
confIntUpper = round(confIntUpper)
```

The efficacy for lgbtq was `r efficLGBTQ`%.  
The 95% Confidence Interval for efficacy was (`r confIntLower`, `r confIntUpper`)%.

### LGBTQ Conclusion

The LGBTQ study was considerably different from the previous two groups. The exact opposite was true for the LGBTQ participants, who experienced significantly fewer illnesses in the placebo group than in the vaccination group. One might take the conclusion that people who are part of the LGBTQ community shouldn't get the vaccine since 73 percent of those who were ill had had vaccinations.


## Druggies

### Graphical Descriptive Results

```{r}
barchartGC(~infected + treatment,data=Druggie)
barchartGC(~infected + treatment,data=Druggie, type="percent")
```

The placebo and vaccine effects on whether or not the drug users were ill are depicted in the left chart. The percentages from the first graph in relation to the section are displayed in the right chart. Similar to the LGBTQ community, those who received the vaccination were roughly 70% more likely to contract the Covid virus than those who received a placebo. Except for LGBTQ, the area "OK" has remained roughly 50/50.

### Numerical Descriptive Results

```{r}
table4 <- xtabs(~infected + treatment, data=Druggie)
rowPerc(table4)
colPerc(table4)
```

This table provides percentages in numbers for patients who received Covid 19. The placebo group represents just around 15% of the observed druggie covid instances, while the vaccination group accounts for 85% of them. This division resembles the LGBTQ subgroup extremely closely.

### Inferrential Results

```{r}
chisq.test(table4)
chisqtestGC(table4)
fisher.test(table4)
```
 
The Fisher exact test gave an odds Ratio of 0.1775019 and a 95% Confidence interval from 0.07654399 to 0.36573284. Since the odds ratio is less than one, the vaccine is .17 times as likley to stop covid as the placebo for Druggies, which is More than the LGBTQ but less than Females. The p-value for the Druggie subgroup is 2.886e-08, which is less than .05, so we must reject the null hypothesis.

### Efficacy

```{r}
efficDruggie = 0.1775019/(0.1775019 + 1) * 100
efficDruggie = round(efficDruggie)
confIntLower = 0.07654399/(0.07654399+ 1) * 100
confIntLower = round(confIntLower)
confIntUpper = 0.36573284/(0.36573284 + 1) * 100
confIntUpper = round(confIntUpper)
```

The efficacy for Druggies was `r efficDruggie`%.  
The 95% Confidence Interval for efficacy was (`r confIntLower`, `r confIntUpper`)%.

### Druggies Conclusion

The Druggie study found that 85% of all documented covid instances originated from the vaccine group, while the remaining 15% originated from the placebo group, similar to the LGBTQ Subgroup.


## Final Discussion and Conclusion

The original question was: "does the vaccine exhibit different effectiveness across the sub populations of Males, Females, LGBTQ members, and habitual intravenous drug users." The data proves that Males have the highest success rate for the vaccine,with females coming in a close second.The vaccine had a highly unfavorable effect on drug users, and LGBTQ people fared the worst in terms of vaccine effectiveness. The sample size for the male and female groups was around six times more than for the druggie and LGBTQ groups. There were 18370 participants in the Females and Males categories, while 3060 people in the Drug Users and LGBTQ categories. The statistics should have an equitable distribution because these are both high populations. The effectiveness of the vaccine was drastically different between the populations. The p-value for all subgroups was less than .05. This means we must reject the null hypothesis of,"The effectiveness of the vaccine does not change between the 4 groups of people."
